Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motions Characteristic to Vertical Axis Turbines

### 2016-08-19 paper.tex
small correction in paragraph
Corrections based on reviewers comments (2)

Static lift and drag coefficients of the 2D profile used for this study are
taken from the technical technical report made by Sheldahl and Klimas
\cite{sheldahl1981aerodynamic}, which is a popular database containing values
for a wide range of Reynolds numbers. These coefficients are tabulated and
combined with the blade element approach for determining body forces acting on
blades. Relative flow velocity $\vec{V}_{\mathrm{rel}}$ and angle of attack
$\alpha$ are calculated for each blade using the geometric relation between the
velocity of the blade $- \omega r$, with $\omega$ representing the angular
(rotational) velocity of the rotor and $r$ the radius of the blade element, and
the velocity of the local incoming wind flow $\vec{V}_{\mathrm{in}}$. Figure
\ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the
plane perpendicular to the turbine axis.

\begin{figure}[h]
%\begin{minipage}{36pc}
\begin{center}
\includegraphics[width=0.35\columnwidth]{vector.eps}
\end{center}
%\resizebox{\columnwidth}{!}{\input{plot}}
%\end{minipage}%
\caption{\label{figvectors} Illustration of velocity vectors and forces acting at the cross-section airfoil element}
\end{figure}

@@

\begin{align*}
\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_d}{2}\right]
\end{align*}

@@

The unsteady attached flow solution considers the circulatory and impulsive loads, which, at the same time depend on the unsteady bound vortex. The circulatory normal force coefficient could be expressed as
\begin{align}
& C_{N_n}^C = C_{N_\alpha} \alpha_{E_n}	\label{CNcirc}
\end{align}
where $C_{N \alpha}$ indicates the slope of the static normal force coefficient for a specific Reynolds number, and $\alpha_{E_n}$ is an expression for an equivalent angle of attack
\begin{align}
& \alpha_{E_n} = \alpha_n - X_n -Y_n - Z_n	\label{alphaeq}
\end{align}
where $\alpha$ is the geometrical angle of attackm $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}. Indices $n$ and $n-1$ in equations \ref{CNcirc} and \ref{alphaeq} represent the current and previous time-step. A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
\begin{align}
& {\alpha _n}' = {\alpha _n} - D_{\alpha _n}    	\label{alphadel}
\end{align}
with $D_\alpha$ as the deficiency function
\begin{align}
& D_{\alpha_n} = D_{\alpha_{n-1}} \exp \left( - \frac{\Delta s}{T_\alpha} \right) + (\alpha_n - \alpha_{n-1})\exp \left( - \frac{\Delta s}{2T_\alpha} \right) \label{defalpha}
\end{align}
with the empirically derived time constant $T_\alpha$, which has a value of $T_\alpha$ = 6.3 for the NACA0021 airfoil, and $\Delta s$ corresponds to a non-dimensional time-step
\begin{align}
& \Delta s = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
\end{align}

Because of the reversal flow in the boundary layer, a leading edge vortex is created at the surface of the airfoil. The critical angle of attack $\alpha_{cr_n}$ defines the condition at which dynamic stall begins
\begin{align}
\alpha_{cr_n} =
\begin{cases}
    \alpha_{ds0}       & \quad |r_n| \geqslant r_0 \\
    \alpha_{ss} + (\alpha_{ds0} - \alpha_{ss}) \frac{| r_n |}{r_0}       & \quad |r_n| < r_0 \\
\end{cases}		\label{criticalpha}
\end{align}
with the reduced pitching rate $r_n$ defined as
\begin{align}
r_n = \frac{\dot{\alpha}_n c}{ 2 | \vec{V}_{rel} | }
\end{align}
here, $\dot{\alpha}$ is the pitch rate, $r_0$ is the reduced pitching rate, which limits the quasi-steady stall with the dynamic stall. Its value is $r_0$ = 0.01 for the NACA0021 airfoil. The static stall onset angle $\alpha_{ss}$ and the critical stall onset angle $\alpha_{ds0}$ are 14.33 and 17.91, respectively. The following is the the dynamic stall condition used
\begin{align}
| \alpha ' | > \alpha_{cr} \rightarrow \mathrm{stall} \label{stall}
\end{align}
for the separated flow, the effects are divided in two groups: trailing edge separation and leading vortex convection. The first one is related to the time delay in the movement of the boundary separation point, and is calculated using Kirchhoff's approximation
\begin{align}
f'_n =
\begin{cases}
    1- 0.4 \exp \left( \frac{|\alpha'_n| - \alpha_1 }{S_1}   \right)    & \quad |\alpha'_n| <  \alpha_1 \\
    0.02- 0.58 \exp \left( \frac{\alpha_1 - |\alpha'_n| }{S_2} \right)  & \quad |\alpha'_n| \geqslant  \alpha_1 \\
\end{cases}		\label{criticalpha2}
\end{align}
where $f'$ is the delayed separation point and $\alpha_1$, $S_1$ and $S_2$ are constant based on the airfoil profile and the local Reynolds number, these can be found in \cite{dyachuk2013dynamic}. The boundary layer around the blade is in function on the time, and this effect is superimposed on the pressure response delay. The additional delay is represented by $f''$, which is the dynamic separation point
\begin{align}
f''_n = f'_n - D _{f_n}	   \label{dyndelay}
\end{align}
and $D_{f_n}$ is the deficiency function defined as
\begin{align}
& D_{f_n} = D_{f_{n-1}} \exp \left( - \frac{\Delta s}{T_f} \right) + (f_n - f_{n-1})\exp \left( - \frac{\Delta s}{2T_f} \right) \label{deff}
\end{align}
Here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
\begin{align}
& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''}}{2} \right)^2	\label{CNf}
\end{align}
after the stall condition is met, the leading edge vortex convects over the surface of the airfoil towards the trailing edge. A significant increase in the normal forces occurs during this process
\begin{align}
C_N^v = B_1 (f'' - f) V_x \label{CNv}
\end{align}
where $C_N^v$ represents the normal force coefficient during the vortex convection, which is in function of the pitch rate. $V_x$ and $B_1$ are parameters based on the local Reynolds number and the airfoil profile, they are found in \cite{dyachuk2013dynamic}. Once the vortex passes the trailing edge, the normal force decreases quickly. Total normal force obtained is expressed as
\begin{align}
C_N = C_{N_n}^f + C_N^v \label{CNtotal}
\end{align}
Figure \ref{figpitching} shows an example of the normal coefficient forces on a blade during pitching motion, using the above model described.

\begin{figure}[h]
%\begin{minipage}{36pc}
\begin{center}
\includegraphics[width=0.5\columnwidth]{CNpitching2.eps}
\end{center}
%\resizebox{\columnwidth}{!}{\input{plot}}
%\end{minipage}%
\caption{\label{figpitching} Illustration of dynamic stall: NACA0021 profile with $\alpha=6 + 16 \sin (\omega	t), \ \omega=12,47$ [rad/s], $V_\infty=34,3$ [m/s] and $c$ = 0,55[m]}
\end{figure}

Tangential force coefficient is calculated trough Kichhoff's flow relation using the dynamic separation point.
\begin{align}
C_T = \eta C_{N_\alpha} \alpha_E ^2 ( \sqrt{f''}-E_0)     \label{CT}
\end{align}
with the empirical constants $\eta = 0.975$ and $E_0 = 0.15$ corresponding to NACA0021 airfoil profile.

\section{Simulation Parameters: Validation case}
A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
reasonable value for operating VAWT) during pitching motion similar to the motion
of the VAWT blade. Different pitching amplitudes were investigated: 13.8, 17.4 and 22.6$\degree$, analog to a TSR of $\lambda=$ 4.19, 3.34 and 2.6 respectively (see equation \ref{tsrval}), to cover a wide range of operational conditions of a VAWT.  In a VAWT with
fixed blades, the angle of attack is in function of the TSR of the turbine. A
periodic function is used to represent the variations of the angles of attack
which a VAWT blade experiences analog to the pitching motion,
\begin{eqnarray}
    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right) \label{alphaval}
\end{eqnarray}
where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
\begin{eqnarray}
    \lambda =  \frac{\Omega R}{ |\vec{V}_\infty | } \label{tsrval}
\end{eqnarray}
In the equation \ref{alphaval}, $\alpha$ represents the geometric angle of attack, and it is different to the effective angle of attack $\alpha_E$, which is calculated by the model in order to get the force coefficients. A NACA0021 profile with 0.55 [m] of chord length is driven to the dynamic stall region trough the pitching motion accordint to equation \ref{alphaval}. The pitching blade experiments were carried out at Glasgow
University\cite{angell1988collected}.

\section{Results and discussion}
In this section, obtained results for $C_N$ and $C_T$ from simulations are compared with the experimental data. Varying parameters have been tested, mainly in the deep stall condition for an amplitude of 22.6 $\degree $ (equivalent to $\lambda$ = 2.6). All the studied cases were simulated using a LES Smagorinsky turbulence model. Figure \ref{revolutions} shows that the model needs two revolutions to reach the convergence, a third revolution will not produce any change.

@@

\textbf{Temporal sensitivity:} Different time-steps were chosen for a varying temporal discretization test, which $\Delta t$ values are 0.001, 0.003, 0.0055 and 0.0075 [s], and these correspond to around 1200, 400, 220 and 160 time-steps per revolution respectively. In this study a coarse mesh with a refinement of 25\%, compared to the reference one, was used. Results are shown in Figure \ref{figdtcoarse}.

@@

According to \cite{dyachuk2015simulating}, where the dynamic stall model is combined with a vortex model, it is beneficial to omit the attached flow correction equation \ref{alphaeq} and instead use $\alpha_E$ = $\alpha_n$. Hence, it is of interest to investigate if this correction should be included or not when the ALM is used instead of the vortex model. Figure \ref{AMP174corrections} depicts that simulated results with the attached flow corrections have small improvements in the accuracy and are closer to experimental data. Results without corrections also have good agreement. Both $C_N$ and $C_T$ peak values are not overestimated.  This same study was carried out for a bigger angle of attack range and differences are less significant.        This has also been tested for larger angles of attack and the differences between using equation \ref{alphaeq} or not are less significant here as the separated flow modeling dominates the force calculation.

>>>>

Static lift and drag coefficients of the 2D profile used for this study are
taken from the technical report made by Sheldahl and Klimas
\cite{sheldahl1981aerodynamic}, which is a popular database containing values
for a wide range of Reynolds numbers. These coefficients are tabulated and
combined with the blade element approach for determining body forces acting on
blades. Relative flow velocity $\vec{V}_{\mathrm{rel}}$ and angle of attack
$\alpha$ are calculated for each blade using the geometric relation between the
velocity of the blade $- \Omega r$, with $\Omega$ representing the angular
(rotational) velocity of the rotor and $r$ the radius of the blade element, and
the velocity of the local incident wind flow $\vec{V}_{\mathrm{in}}$ which is generally lower than the incoming flow velocity $\vec{V}_{\infty}$. Figure
\ref{figvectors} depicts a cross-sectional airfoil element at radius $r$ in the
plane perpendicular to the turbine axis.

\begin{figure}[h]
%\begin{minipage}{36pc}
\begin{center}
\includegraphics[width=0.33\columnwidth]{vector2.eps}
\end{center}
%\resizebox{\columnwidth}{!}{\input{plot}}
%\end{minipage}%
\caption{\label{figvectors} Illustration of velocity vectors and forces acting at the cross-section airfoil element}
\end{figure}

@@

\begin{align*}
\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_D}{2}\right]
\end{align*}

@@

The unsteady attached flow solution considers the circulatory and impulsive loads, which, at the same time depend on the unsteady bound vortex. The circulatory normal force coefficient could be expressed as
\begin{align}
& C_{N_n}^C = C_{N_\alpha} \alpha_{E_n}	\label{CNcirc}
\end{align}
where $C_{N \alpha}$ indicates the slope of the static normal force coefficient for a specific Reynolds number, and $\alpha_{E_n}$ is an expression for an equivalent angle of attack
\begin{align}
& \alpha_{E_n} = \alpha_n - X_n -Y_n - Z_n	\label{alphaeq}
\end{align}
where $\alpha$ is the geometrical angle of attack and $X$, $Y$ and $Z$ are the deficiency functions, which are empirically derived, based on the flow velocity and the pithing rate, the details are found in \cite{dyachuk2013dynamic}. Indices $n$ and $n-1$ represent the current and previous time-step. A delayed angle of attack is considered due to the lag in pressure response, and it is calculated as
\begin{align}
& {\alpha _n}' = {\alpha _n} - D_{\alpha _n}    	\label{alphadel}
\end{align}
with $D_\alpha$ as the deficiency function
\begin{align}
& D_{\alpha_n} = D_{\alpha_{n-1}} \exp \left( - \frac{\Delta s}{T_\alpha} \right) + (\alpha_n - \alpha_{n-1})\exp \left( - \frac{\Delta s}{2T_\alpha} \right) \label{defalpha}
\end{align}
with the empirically derived time constant $T_\alpha$, which has a value of $T_\alpha$ = 6.3 for the NACA0021 airfoil, and $\Delta s$ corresponds to a non-dimensional time-step
\begin{align}
& \Delta s = \frac{2 | \vec{V}_{rel} | \Delta t}{c}		.	\label{deltas}
\end{align}

Because of the reversal flow in the boundary layer, a leading edge vortex is created at the surface of the airfoil. The critical angle of attack $\alpha_{cr_n}$ defines the condition at which dynamic stall begins
\begin{align}
\alpha_{cr_n} =
\begin{cases}
    \alpha_{ds0}       & \quad |r_n| \geqslant r_0 \\
    \alpha_{ss} + (\alpha_{ds0} - \alpha_{ss}) \frac{| r_n |}{r_0}       & \quad |r_n| < r_0 \\
\end{cases}		\label{criticalpha}
\end{align}
with the reduced pitching rate $r_n$ defined as
\begin{align}
r_n = \frac{\dot{\alpha}_n c}{ 2 | \vec{V}_{rel} | }
\end{align}
here, $\dot{\alpha}$ is the pitch rate, $r_0$ is the reduced pitching rate, which limits the quasi-steady stall with the dynamic stall. Its value is $r_0$ = 0.01 for the NACA0021 airfoil. The static stall onset angle $\alpha_{ss}$ and the critical stall onset angle $\alpha_{ds0}$ are 14.33 and 17.91, respectively. The following is the the dynamic stall condition used
\begin{align}
| \alpha ' | > \alpha_{cr} \rightarrow \mathrm{stall} \label{stall}
\end{align}
for the separated flow, the effects are divided in two groups: trailing edge separation and leading vortex convection. The first one is related to the time delay in the movement of the boundary separation point, and is calculated using Kirchhoff's approximation
\begin{align}
f'_n =
\begin{cases}
    1- 0.4 \exp \left( \frac{|\alpha'_n| - \alpha_1 }{S_1}   \right)    & \quad |\alpha'_n| <  \alpha_1 \\
    0.02- 0.58 \exp \left( \frac{\alpha_1 - |\alpha'_n| }{S_2} \right)  & \quad |\alpha'_n| \geqslant  \alpha_1 \\
\end{cases}		\label{criticalpha2}
\end{align}
where $f'$ is the delayed separation point and $\alpha_1$, $S_1$ and $S_2$ are constant based on the airfoil profile and the local Reynolds number, these can be found in \cite{dyachuk2013dynamic}. The boundary layer around the blade is in function on the time, and this effect is superimposed on the pressure response delay. The additional delay is represented by $f''$, which is the dynamic separation point
\begin{align}
f''_n = f'_n - D _{f_n}	   \label{dyndelay}
\end{align}
and $D_{f_n}$ is the deficiency function defined as
\begin{align}
& D_{f_n} = D_{f_{n-1}} \exp \left( - \frac{\Delta s}{T_f} \right) + (f_n - f_{n-1})\exp \left( - \frac{\Delta s}{2T_f} \right) \label{deff}
\end{align}
Here, $T_f$ is empirically derived time constant and its value is $T_f$ = 3. Then, the normal force coefficient for unsteady conditions before the dynamic stall onset is calculated as
\begin{align}
& C_{N_n}^f = C_{N_\alpha} \alpha_{E_n} \left( \frac{1 + \sqrt{f''_n}}{2} \right)^2	\label{CNf}
\end{align}
after the stall condition is met, the leading edge vortex convects over the surface of the airfoil towards the trailing edge. A significant increase in the normal forces occurs during this process
\begin{align}
C_N^v = B_1 (f''_n - f_n) V_x \label{CNv}
\end{align}
where $C_N^v$ represents the normal force coefficient during the vortex convection, which is in function of the pitch rate. $V_x$ and $B_1$ are parameters based on the local Reynolds number and the airfoil profile, they are found in \cite{dyachuk2013dynamic}. Once the vortex passes the trailing edge, the normal force decreases quickly. Total normal force obtained is expressed as
\begin{align}
C_N = C_{N_n}^f + C_N^v \label{CNtotal}
\end{align}
Figure \ref{figpitching} shows an example of the normal coefficient forces on a blade during pitching motion, using the above model described.

\begin{figure}[h]
%\begin{minipage}{36pc}
\begin{center}
\includegraphics[width=0.5\columnwidth]{CNpitching2.eps}
\end{center}
%\resizebox{\columnwidth}{!}{\input{plot}}
%\end{minipage}%
\caption{\label{figpitching} Illustration of dynamic stall: NACA0021 profile with $\alpha=6 + 16 \sin (\Omega	t), \ \Omega=12,47$[rad/s], $V_\infty=34,3$[m/s] and $c$ = 0,55[m]}
\end{figure}

Tangential force coefficient is calculated trough Kichhoff's flow relation using the dynamic separation point.
\begin{align}
C_T = \eta C_{N_\alpha} \alpha_E ^2 ( \sqrt{f''_n}-E_0)     \label{CT}
\end{align}
with the empirical constants $\eta = 0.975$ and $E_0 = 0.15$ corresponding to NACA0021 airfoil profile.

\section{Simulation Parameters: Validation case}
A NACA0021 airfoil was tested at the Reynolds number of 1.000.000 (which is a
reasonable value for operating VAWT) during pitching motion similar to the motion
of the VAWT blade. Different pitching amplitudes were investigated: 13.8, 17.4 and 22.6$\degree$, analog to a TSR of $\lambda=$ 4.19, 3.34 and 2.6 respectively (see equation \ref{tsrval}), to cover a wide range of operational conditions of a VAWT.  In a VAWT with
fixed blades, the angle of attack is in function of the TSR of the turbine. A
periodic function is used to represent the variations of the angles of attack
which a VAWT blade experiences analog to the pitching motion,
\begin{eqnarray}
    \alpha = \arctan \left( \frac{\sin \theta}{\lambda + \cos \theta} \right) \label{alphaval}
\end{eqnarray}
where $\theta$ is the azimuthal blade angle and $ \lambda $ represents the TSR,
\begin{eqnarray}
    \lambda =  \frac{\Omega R}{ |\vec{V}_\infty | } \label{tsrval}
\end{eqnarray}
In the equation \ref{alphaval}, $\alpha$ represents the geometric angle of attack, and it is different to the effective angle of attack $\alpha_E$, which is calculated by the model in order to get the force coefficients. A NACA0021 profile with 0.55[m] of chord length is driven to the dynamic stall region trough the pitching motion accordint to equation \ref{alphaval}. The pitching blade experiments were carried out at Glasgow
University\cite{angell1988collected}.

\section{Results and discussion}
In this section, obtained results for $C_N$ and $C_T$ from simulations are compared with the experimental data. Varying parameters have been tested, mainly in the deep stall condition for an amplitude of 22.6 $\degree $ (equivalent to $\lambda$ = 2.6). All the studied cases were simulated with a domain geometrically similar to the wind tunnel used in \cite{angell1988collected} to emulate the pitching blade experiments: 1.61[m] of span, 2.13[m] of height, 2.5[m] upstream and 4[m] downstream, with similar velocity boundary conditions: 28.4[m/s] in the inlet and no-slip condition (fixed 0[m/s]) in the walls, using a LES Smagorinsky turbulence model. Figure \ref{revolutions} shows that the model needs two revolutions to reach the convergence, a third revolution will not produce any change.

@@

\textbf{Temporal sensitivity:} Different time-steps were chosen for a varying temporal discretization test, which $\Delta t$ values are 0.001, 0.003, 0.0055 and 0.0075[s], and these correspond to around 1200, 400, 220 and 160 time-steps per revolution respectively. In this study a coarse mesh with a refinement of 25\%, compared to the reference one, was used. Results are shown in Figure \ref{figdtcoarse}.

@@

According to \cite{dyachuk2015simulating}, where the dynamic stall model is combined with a vortex model, it is beneficial to omit the attached flow correction equation \ref{alphaeq} and instead use $\alpha_E$ = $\alpha_n$. Hence, it is of interest to investigate if this correction should be included or not when the ALM is used instead of the vortex model. Figure \ref{AMP174corrections} depicts that simulated results with the attached flow corrections have small improvements in the accuracy and are closer to experimental data. Results without corrections also have good agreement. Both $C_N$ and $C_T$ peak values are not overestimated. This has also been tested for larger angles of attack and the differences between using equation \ref{alphaeq} or not are less significant here as the separated flow modeling dominates the force calculation.
